---
title: "Lab1 - Esther Xuanpei Ouyang"
author: "Esther Xuanpei Ouyang"
date: "1/22/2018"
output:
  html_document: default
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## 1) Basic Vector and Matrix manipulations in R

```{r}
x = 1:9

matrix(data = x, nrow = 3, ncol = 3, byrow = FALSE)

matrix(data = x, nrow = 3, ncol = 3, byrow = TRUE)

diag(x = 1, 5, 5)

a1 <- c(2, 3, 6, 7, 10)
a2 <- c(1.88, 2.05, 1.70, 1.60, 1.78)
a3 <- c(80, 90, 70, 50, 75)

A = cbind(a1, a2, a3)
A

b1 <- c(1, 4, 5, 8, 9)
b2 <- c(1.22, 1.05, 3.60, 0.40, 2.54) 
b3 <- c(20, 40, 30, 80, 100)

B = rbind(b1, b2, b3)
B
A%*%B
B%*%A
t(A)%*%t(B)
t(B)%*%t(A)


head(iris)

selected_iris = cbind(iris[,1],iris[,2],iris[,3],iris[,4])
parameters = as.matrix(c(1,2,3,4),4,1)
lincombo = selected_iris%*%parameters
lincombo

v = 1:5
vnorm = function(x) {
  return(sqrt(t(x)%*%x))
}
vnorm(v)
unit_u = v / vnorm(v)
unit_u

is_square = function(x) {
  dimen = dim(x)
  return (dimen[1] == dimen[2])
}

x = 1:8
test1 = matrix(x, 2, 4)
test1
is_square(test1)

y = 1:9
test2 = matrix(y, 3, 3)
test2
is_square(test2)

mtrace = function(mat) {
  if (!is_square(mat)) {
    return (0)
  } else {
    dimen = dim(mat)
    I = diag(1, dimen[1], dimen[2])
    return(sum(sum(mat*I)))
  }
}

test1 = matrix(1:9, 3, 3)
test2 = matrix(2:10, 3, 3)
test3 = test1 + test2
# Test 1
mtrace(test1) + mtrace(test2)
mtrace(test3)

# Test 2
test4 = 3 * test1
mtrace(test1) * 3
mtrace(test4)

# Test3
X = test1
Y = test2
XtY = t(X)%*%Y
XYt = X%*%t(Y)
YtX = t(Y)%*%X
YXt = Y%*%t(X)

mtrace(XtY)
mtrace(XYt)
mtrace(YtX)
mtrace(YXt)
```

#### Proof for Trace

##### 1. $tr(A+B)=tr(A)+tr(B)$

Suppose $A$, $B$ are both square matrix with dimention n x n.
Let $C=A+B$, then from the definition of matrix addition, $c_{ij} = a_{ij}+b_{ij}$.
By the definition of trace, $$tr(A)+tr(B) \\ =\sum_{i=1}^n a_{ii}+\sum_{i=1}^n b_{ii} \\ =\sum_{i=1}^n a_{ii} + b_{ii} = \sum_{i=1}^nc_ii \text{  (definition of matrix addition)} \\ = tr(C)$$

##### 2. $tr(cA) = c \times tr(A)$, where c is a scalar.
Suppose $A$, $B$ are both square matrix with dimention n x n.
Let $D=cA$, then from the definition of scalar matrix multiplication, $d_{ij} = c\times a_{ij}$.
By the definition of trace, $$tr(cA) \\ =\sum_{i=1}^n c\times a_{ii} \\ =c \times \sum_{i=1}^n a_{ii} \text{  (take c out of the summation form since c is a scalar)} \\ = c \times tr(A)$$

##### 3. $tr(X^TY) = tr(XY^T) = tr(Y^TX) = tr(YX^T)$

Suppose $X$, $Y$ are both square matrix with dimension n x n, $X = \begin{bmatrix} x_{11} & \dots & x_{1n} \\ \vdots & \ddots & \vdots \\ x_{n1} & \dots & x_{nn}  \end{bmatrix} = \begin{bmatrix}  -  x_1 -\\  \vdots  \\  -x_{n}-  \end{bmatrix}, \text{where } x_i \text{ is the i row vector of X}$, $Y = \begin{bmatrix} \vert & & \vert \\ y_1  & \dots & v_2   \\ \vert & & \vert \end{bmatrix}, \text{where } y_i \text{ is the i column vector of Y}$

First, we prove that $(AB)^T = B^TA^T$ by $$((AB)^T)_{ij} = (AB)_{ji} = \sum_k A_{jk}B_{ki} = \sum_k(A^T)_{kj}(B^T)_{ik} = \sum_k(B^T)_ik (A^T)_{kj} = (B^TA^T)_{ij} \text{(1)}$$
Then, we prove that $tr(A) = tr(A^T)$ by $$tr(A) = \sum_{i=1}^n a_{ii} = tr(A)^T  \text{since tranpose does not change the numbers on the diagonal values } a_{ii} \text{(2)}$$

Last, suppose $C$,$D$ are both sqaure matrix with dimension n x n, we prove that $tr(CD) = tr(DC)$ by 
By definition of matrix-matrix multiplication, $A_{ij}=x_i^Ty_j$, which means the inner product of the row-vector $x_i$ and the column-vector $y_j$.

Therefore, 
$$tr(CD)=tr(E) =\sum_{i=1}^nE_{ii} = \sum_{i=1}^nc_{i}^Td_{i}=\sum_{i=1}^nd_{i}^Tc_{i}=\sum_{i=1}^nF_{ii}=tr(F)=tr(DC) \text{ (3)}$$
Let $A = X^TY$, by (1) and (2), $tr(A) = tr(A^T) \Rightarrow tr(X^TY) = tr((X^TY)^T) = tr(Y^TX)$.

Let $B = XY^T$, by (1) and (2), $tr(B) = tr(B^T) \Rightarrow tr(XY^T) = tr((XY^T)^T) = tr(YX^T)$.

By (3), $tr(X^TY) = tr(YX^T)$ and $tr(XY^T) = tr(Y^TX)$.

Therefore,  $tr(X^TY) = tr(XY^T) = tr(Y^TX) = tr(YX^T)$.


## 2) Transformation and Scaling Operations

```{r}
head(mtcars)

# Create a matrix M with variables mpg, disp, hp, drat, and wt.
M = mtcars[, c(1,3,4,5,6)]
M

# Use apply() to compute the vector containing the means of the columns in M
apply(M, 2, mean)

# Compute a matrix Mc of mean-centered data applying the function scale() on M (do NOT use the argument scale = TRUE).
Mc = scale(M, scale = FALSE)
Mc
# Confirm that variables in Mc are mean-centered by calculating the vector of column-means 
apply(Mc, 2, mean)

# Use the function sweep() to mean-center M by sweeping out the vector of column means.
M.mean = apply(M, 2, mean)
M.mean
Msweep = sweep(M, 2, M.mean)
Msweep

# Compare this result with Mc (you should get the same values).
all(Mc == Msweep)

# Compute a vector of column maxima from M.
Mmax = apply(M, 2, max)
Mmax

# Use sweep() to scale the columns of M by dividing by the column maxima.
sweep(M, 2, Mmax,"/")

# Compute a matrix in which all columns of M are scaled such that they have minimum = 0 and maximum = 1
Mmax = apply(M, 2, max)
Mmin = apply(M, 2, min)
Mrange = Mmax - Mmin
scaled_M = scale(M , center = Mmin, scale = Mrange)
scaled_M
apply(scaled_M, 2, min)
apply(scaled_M, 2, max)

# Without using the function cov(), compute the sample covariance matrix of the variables in M: mpg, disp, hp, drat, and wt.
n = dim(mtcars)[1]
Mcov = (t(Mc)%*%Mc)/(n-1)
Mcov

# Without using the function cor(), compute the correlation matrix of the variables in M: mpg, disp, hp, drat, and wt.
standardize_M = scale(M) # variance = 1
Mcor = (t(standardize_M)%*%standardize_M)/(n-1)
Mcor
```

```{r}
cyl = factor(mtcars$cyl)

dummify = function(char_vector, all) {
  n_class = nlevels(char_vector)
  if(!all) {
    res = matrix(0,1,n_class-1)
  } else {
    res = matrix(0,1,n_class)
  }
  dummy_mat = diag(1, n_class, n_class)
  if (!all) {
    dummy_mat = dummy_mat[, -1]
  }
  class_level = c(levels(char_vector))
  for (class in char_vector) {
    class_i = match(class, class_level)
    res = rbind(res, dummy_mat[class_i, ])
  }
  return(res[-1,])
}

dummify(cyl, all = TRUE)
dummify(cyl, all = FALSE)

gear = factor(mtcars$gear)

crosstable = function(vec1, vec2) {
  dummy1 = dummify(vec1, all = TRUE)
  dummy2 = dummify(vec2, all = TRUE)
  return(t(dummy1)%*%dummy2)
}

xtb <- crosstable(cyl, gear)
xtb
```

